---
title: "MonthlyTimeSeriesEnergyUS"
author: "Logan Sowards"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
GOAL:
Read in US Energy Consumption data to forecast future energy consumption

READ IN DATA:
  US Residential Energy Consumption Since 1991
  Data re-downloaded 2/24/21 (Original Code from 1/15/20)
  https://www.eia.gov/totalenergy/data/browser/?tbl=T02.02#/?f=M
  http://www.eia.gov/totalenergy/data/monthly/index.cfm#consumption
```{r}
data1 <- read.csv("http://www.eia.gov/totalenergy/data/browser/csv.cfm?tbl=T02.01")

# subset to TERCBUS Total Energy Consumed by the Residential Sector
data2 <- subset(data1, MSN == "TERCBUS")

# subset to "your lifetime"
data3 <- subset(data2, data2$YYYYMM > 199100)

# remove yearly total (coded "month 13", every 13th obs)
data4 <- subset(data3, data3$YYYYMM %% 100 != 13)

energy <- data4$Value
```


EDA:
```{r}
plot(energy,  type = "b", main = "US Residential Energy Consumption",
     ylab = "Energy Consumption (trillion Btu)", xlab = "Months Since January 1991")

# Notice that in addition to a slight increasing trend, there is also a seasonal trend (12 month cycle)
tail(energy)
```



VISUALIZE MONTHLY PATTERN:
```{r}
plot(1:12,energy[1:12],
     ylim=c(1000,3000),
     type="b",
     main = "US Residential Energy Consumption by Month",
     ylab="US Residential Energy (trillion Btu)",
     xlab="Month")
for(i in 1:27){
  lines(1:12,energy[12*i + (1:12)],type="b")
}
```



ANALYSIS:
```{r}
#Model: ARIMA (1,1,1) x (1,1,1)_12
library(astsa)
energy.out <- sarima(energy, 1,1,1, 1,1,1, 12)

energy.out$ttable

```


PREDICTION:
```{r}
#Predict Energy Usage up to end of 2022

# determine where data ends
tail(data4)
#data ends in November 2020

future.month <- c(12, rep(1:12, 2))
future.year <- c(2020, rep(2021:2022, each = 12))

# Predictions
energy.future <- sarima.for(energy, 1,1,1, 1,1,1, 12, n.ahead = length(future.year))
```


TABLE OF PREDICTION AND CONFIDENCE INTERVALS:
```{r}
# Compute 95% prediction intervals
L = energy.future$pred - 2*energy.future$se
U = energy.future$pred + 2*energy.future$se

# Table of predictions
cbind(future.month, future.year, energy.future$pred, L, U)
```


ANALYSIS:

Research Task: Prediction (2 years in the future)

Data Features: Seasonal Time Series (Monthly Data)

Strengths: Incorporated Seasonal Pattern
          More Frequent Data
          Predict with Prediction Intervals
          Prediction matches intuition
          long and short memory


Weaknesses: We know seasonality (We need to know in order for model to make sense)
            "Overall trend" is really "every month has a trend"
            Graph doesn't suggest the reasons for the trend
            Assumes past predicts future
