---
title: "RandomForestSBA"
author: "Logan Sowards"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Tree Model HW
Reproducable Code
Data: California SBA Loans
Data Source: https://www.tandfonline.com/doi/suppl/10.1080/10691898.2018.1434342/suppl_file/ujse_a_1434342_sm1547.zip

READ DATA:
```{r}
temp <- tempfile()
download.file("https://www.tandfonline.com/doi/suppl/10.1080/10691898.2018.1434342/suppl_file/ujse_a_1434342_sm1547.zip",temp)
default <- read.csv(unz(temp, "SBAcase.11.13.17.csv"))
unlink(temp)

#find names and indexes of columns
names(default)
```






FILTER DATA:
```{r}
# Make Dataset more useful by removing columns: ( MIS_Status[25], ChgOffDate[21] and
# ChgOffPrinGr[26], Selected[1], Name[3], Recession[32], daysterm[33], City[4], State[5],
# Bank[7], xx[34])
SBAfilter <- default[,-c(1,3,4,5,7,21,25,26,32,33,34)]

# Check if proper columns removed
names(SBAfilter)


# Feature Engineering: Add columns for first 2 and 3 digits from Zipcode,
# first 3 and 4 digits from NAICS code (check for more specific explanatory variables)
SBAfilter$Zip2Digit  <- floor(SBAfilter$Zip / 1000) # Dividing by 1000 makes to a decimal and floor removes the decimals
SBAfilter$Zip3Digit  <- floor(SBAfilter$Zip / 100)
SBAfilter$NAICS3digit <- floor(SBAfilter$NAICS / 1000)
SBAfilter$NAICS4digit <- floor(SBAfilter$NAICS / 100)


# Filter out rows w/ any values of NewExist that are not 1 or 2
# and any missing values of DisbursementDate
SBAfilter <- SBAfilter[c(SBAfilter$NewExist == 1 | SBAfilter$NewExist == 2), ] # | is the or operand
SBAfilter <- SBAfilter[!is.na(SBAfilter$DisbursementDate), ] #!is.na() returns areas that are not empty

# Check if Default is a factor and if not, convert to a factor
is.factor(SBAfilter$Default)
SBAfilter$Default <- factor(SBAfilter$Default)
is.factor(SBAfilter$Default)

# Show the dimensions of the data to show it is correct
dim(SBAfilter)
table(SBAfilter$Default)
```



IDENTIFY VARIABLES:

Default (if the loan was paid or charged off) is the response variable. It is categorical.

Quantitative Explanatory Variables: LoanNr_ChkDgt, Zip, Zip2Digit, Zip3Digit, NAICS, NAICS3digit, 
NAICS4digit, Term, NoEmp, CreateJob, RetainedJob, DisbursementGross, BalanceGross, GrAppv,
SBA_Appv, Portion

Categorical Explanatory Variables: BankState, ApprovalDate, ApprovalFY, NewExist, FranchiseCode, 
UrbanRurual, RevLineCr, LowDoc, DisbursementDate, New, RealEstate



CREATE MODEL DATASET:
```{r}
set.seed(20)

# construct model dataset (all bads, sample of goods b/c bads are rarer than good in this set)
# bad = default(1), goods = paid(0)

# Collect all bads
all.bad <- subset(SBAfilter, Default == 1) 
n.bads <- dim(all.bad)[1]

# SRS of goods
all.good <- subset(SBAfilter, Default == 0)
n.goods <- dim(all.good)[1]
rows.good <- sample(n.goods, n.bads)
sample.good <- all.good[rows.good,]

#put goods and bads together
loan.model <- rbind(all.bad, sample.good)


# Test that dimensions of model are correct
dim(loan.model)

# confirm loan.model has half goods and half bads
table(loan.model$Default)
```


CREATE TRAIN AND TEST DATASETS:
```{r}
n.model <- dim(loan.model)[1]

# 70% train, 30% test good rule of thumb
0.7*n.model #960.4
train.rows <- sample(n.model, 960)
loan.train <- loan.model[train.rows,]

loan.test <- loan.model[-train.rows,]

# confirm train and test are similar
table(loan.train$Default)
table(loan.test$Default)
```


RANDOM FOREST MODEL:
```{r}
# random forest library
library(randomForest)

# Make a prediction of a future obs

names(loan.train) # Check which column the response is in so you can take it out of the tree 


loan.out <- randomForest(x = loan.train[,-24], y=loan.train$Default, # use -24 to take out the response variable which is in -24
                           xtest = loan.test[,-24], ytest = loan.test$Default,
                           keep.forest=TRUE, # store trees for future prediction
                           replace = TRUE, # use bootstrap sampling (SRS)
                           ntree = 50, # number of trees (50 is generally a good number)
                           mtry = 10, # lets other numbers have a shot to be a tree. Rule of thumb is p/3 (p = number of explanatory variables, round down)
                           nodesize = 20) #how many in each bin minimum, needs to be biggger in not stratifying 

loan.out # Want MSE and % Var Explained to be about the same for each, we also want low MSE
# Error rate: How many times were we wrong
# Want test and train error rate to be about the same. 

```


PLOTS:
```{r}
#shows the use of trees in prediction
plot(loan.out)
```


```{r}
# Does the model make sense? Intuition of what matters in the model
varImpPlot(loan.out)
```


SOME CONCLUSIONS:
Term, ApprovalDate, and SBA_Appv appear to be the three most influential values

This model neither confirms nor refutes that urban loans are more risky. 
It suggests that Urban vs Rural loans may slightly help predict if a loan will be prepaid
However it does not suggest which one is more risky.
